\papertitle

### 2020-01-28 paper/paper.tex
Cleanup of old stuff

@@

@@

Figure \ref{fig:cycle-4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
Assume you have, for instance, four cycles of the model running with all the required post-processing steps.
Typically, the user has generated the scripts for each task by generating a filename using the same prefix that is then placed at the same storage type every time.
That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.

The user-specific storage locations are a viable option to run a workflow.
The user might even have analysed various scenarios such as depicted in
\Cref{sec:datamanagement}.
However, by automating these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.

@@

\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics.
This will increase performance and system utilization.

\item Performance-portability: The user has to specify the IO requirements only once for the tasks of the workflow.
The workflow can now be run on any system without user intervention.
Even more, if the system characteristics change, e.g.. it gets upgraded and an additional storage tier becomes available or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
\end{itemize}

\section{Design}
What changes now in terms of specification?

In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the datasets/tasks.

\item The user will insert this information as an additional file to the Cylc workflow management. In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible. Here, we are thinking about simplifying the information the user is already providing. So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:

@@

\item The second option to access this type of information from the user does not require any direct information at all. The idea here is to use the DDN IME Monitoring system and have an estimate for the parameters after one run of the complete workflow. This approach is beneficial to the user because s/he does not have to take any extra step that s/he is currently doing to run the workflow through Cylc. However, the price here is being paid with an additional run of the complete workflow, which can be demanding at both time and expenses (we are considering here the paid access to a supercomputer, for example). In the end, it is up to the user and the previous knowledge s/he has about the intermediate stages of the workflow.

@@

\section{Extending Existing Workflows}

The focus of this paper is to exemplify the I/O optimisations that can be made if the datasets dependencies are established. This information, in fact, is presented in the workflows before the Cylc application, so nothing else will be required from the user. But to have the information explicitly, Cylc needs some changes to be able to:

\begin{enumerate}

\item deliver it together with the workflow, and

\item adapts the resulting workflow to consider the I/O modifications.

\end{enumerate}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\columnwidth]{cycle1}
  \caption{First stage workflow}
  \label{fig:cycle1}
\end{figure}

The scenario presented in Figure \ref{fig:cycle-io} shows the same cycle of Figure \ref{fig:cycle1}, but the extended information about datasets is introduced.
In this picture, we have the tasks dependencies, and we explicit the datasets dependencies regarding the tasks that generated them.
The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination. The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool. The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{cycle-io}
  \caption{First stage workflow including datasets}
  \label{fig:cycle-io}
\end{figure}

Figure \ref{fig:cycle-io-dep} now introduces one example of dependencies among the datasets and tasks: datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle. The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one. Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}. That is not an immediate conclusion when we analyse just the arrows for the dependencies. It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}. Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later. From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow. This information today is static and manual, leaving opportunities for automatisation and optimisations open.
The idea here is to embrace the concept that tasks dependencies are really datasets dependencies. Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{cycle-io-dep}
  \caption{First stage workflow including datasets and dependencies}
  \label{fig:cycle-io-dep}
\end{figure}

@@

  \item \textbf{Application}\ \ The user-specified parallel application runs.
  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
  Knowing the potential targets, ESDM schedules application-local I/O.
  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.

\end{enumerate}

@@

@@

\section{Challenges and Limitations}
\label{sec:challenges}
The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.

\begin{description}
\item[Explicit Workflows]

Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and \sout{{\color{blue}{output}}} {\color{cyan}{products}}.
The computing-intense parts ({\color{blue}{jobs}} {\color{cyan}{tasks?}}) and their dependencies are administrated by a cluster resource manager that dispatches them.
These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.

\item[Suboptimal Storage Tiering]

Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations. On top of that, it is a tedious and error-prone task.

Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) on a particular storage system and, usually, moving data between storage by using scripts.
One limitation here is that decisions about how data structures are mapped and packaged into \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) are made once by the producing application, and cannot be changed without manual intervention by a downstream application.

A policy system (e.g. burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.

Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
In any case, data migration needs resources from the source {\color{cyan}{source? the machine?}} and the target storage system to copy data.
Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling \sout{{\color{blue}{the data source producing data with a data consumer}}} {\color{cyan}{the data source with the data consumer}}) to prevent storing intermediate products on persistent storage.

\item[Task and Data Equality]

All tasks and data products are handled identically by a HPC system, meaning there is no differentiation between individual jobs and data products.
This lack of hierarchy is problematic, as different use cases may not only differ in priority and deadline, but the computing speed usually depends on the critical path of the execution.

This problem is even more severe for data products: a storage system provides a single level of performance and resilience \sout{{\color{blue}{and performance}}} for stored files.
\sout{{\color{blue}{files}}} {\color{cyan}{datasets}})
Theoretically, different fault-tolerance and performance policies can be realised with varying storage systems and using management policies.
However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files. {\color{cyan}{I don't think we will be dealing with that in the next 10 years!}}

\item[Manual Optimisation]

Utilising the computing and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging.
Even within a single data centre, utilising homogeneous storage and computing infrastructure efficiently is complex for experts and needs manual definition.
The efficient management of data and computing capabilities in a heterogeneous environment and
how the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures
are still unresolved questions.

The optimisation of a single parallel job already is quite challenging.
To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
Similarly, efficient data placement is complex. But a specific workflow combined with manual optimisation is the root-cause of subsequent problems, and it must be avoided and replaced with automated and optimised decisions.

Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.

\section{Vision}

Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.

\begin{description}
\item[Workflow Specification]

The user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
The definition of constraints, for instance, deadlines for the generation of the final data products (insight), can also be included.
Workflows can be defined using arbitrary tools, including Python.
Note that a user will not provide an accurate mapping to system or infrastructure, but expects that the system derives an optimised execution plan that covers heterogeneous infrastructures.
It does not matter where data is processed and how and where information is stored, but only that the desired result is computed with all constraints met.
Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing reproduction and validation of the correct execution of experiments.

\item[Smart Scheduling]

Supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources efficiently.
As the API is in charge of both tasks, it may decide to trade computation and storage capacity considering characteristics as costs and energy consumption, for example. Additionally, it can enable automatic recomputation of intermediate states utilising virtualisation and container technologies.
Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.

Automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
Note that this is a significant change for HPC environments, as computing resources are typically dedicated to specific jobs.

@@

\item[Input] The information inserted in the workflow.

\item[Tasks] The tasks specified by the user.

\item[Products] Any of the possible outcomes of the tasks.

For simplicity, we will call these datasets as \texttt{product-workflow.ds}. In this specific example, it is composed by the datasets \texttt{product1.ds}, \texttt{product2.ds}, and \texttt{product3.ds}.

Now.

\begin{enumerate}

  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.

  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.

  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.

  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.

  \item \textbf{Script}\ \ The shell script executes the users-provided instructions allowing Cylc to generate its internal variables. To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.

  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.

\end{enumerate}

Our Plan.

>>>>

Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system and, usually, moving data between storage by using scripts.
One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.

A policy system (e.g., deployed on a burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.

@@

Data system do not respect the importance of data depends at all; aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow are not considered.
Value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability.

@@

Figure \ref{fig:cycle-4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
Assume you have, for instance, four cycles of the model running with all the required post-processing steps.
Typically, the user has generated the scripts for each task by generating a filename using the same prefix that is then placed at the same storage type every time.
That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
The user-specific storage locations are a viable option to run a workflow.
The user might even have analysed various scenarios such as depicted in
\Cref{sec:datamanagement}.

Additionally, by knowing how to rerun the workflow behind the data creation, a smarter storage system could trade data security with potential recomputation opportunities to optimize the cost-efficiency of the system.
Intermediate states could be rerun by utilising virtualisation and container technologies.
Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
Finally, data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.

Ultimately, by abstracting and automatizing these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.

@@

\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics and storage resilience.
This will increase performance and system utilization.
By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimize the cost-efficiency.

\item Performance-portability: The user has to specify the IO requirements only once for the tasks of the workflow.
The workflow can now be run on any system without user intervention.
Even more, if the system characteristics change, e.g.. it gets upgraded and an additional storage tier becomes available or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.

\section{Design}

This section describes our approach to incrementally extent the current software stack for climate and weather workflows that realizes parts our vision.

In order to automatically make scheduling decisions the software stack needs to:
\begin{enumerate}
\item deliver information about information life cycle together with the workflow, and
\item adapt the resulting workflow, individual script and application execution to consider the automatic IO schedule.
\end{enumerate}

\subsection{Transiting to Data Dependencies}

\begin{figure}[bt]
  \begin{subfigure}{0.45\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{cycle1}
    \caption{Original}
    \label{fig:cycle1}
  \end{subfigure}
  \qquad
  \begin{subfigure}{0.45\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{cycle-io-dep}
    \caption{Including datasets and dependencies}
    \label{fig:cycle-io-dep}
  \end{subfigure}

  \caption{Simplified Cylc Workflow for the first stage}
\end{figure}

The scenario presented in \Cref{fig:cycle-io-dep} shows the same cycle of Figure \ref{fig:cycle1}, but  extended information about datasets is added.
In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool. The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.

Datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle. The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one. Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}. That is not an immediate conclusion when we analyse just the arrows for the dependencies. It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}. Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later. From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow. This information today is static and manual, leaving opportunities for automatisation and optimisations open.
The idea here is to embrace the concept that tasks dependencies are really datasets dependencies. Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.

The user to be able to make decisions about storage and the processing of the datasets/tasks.
user will insert this information as an additional file to the Cylc workflow management.
In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible. Here, we are thinking about simplifying the information the user is already providing. So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:

@@

\subsection{Smarter Scheduling}

@@

@@

  \item \textbf{Application}\ \ The user-specified parallel application runs.
  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
  Knowing the potential targets, ESDM schedules application-local I/O.
  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.
\end{enumerate}

@@

Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
%On top of that, it is a tedious, non-portable, and error-prone task.
%Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.

@@

@@
